---
title: "Assignment 2: Coding Basics"
author: "Stephanie Kinser"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your first and last name into the file name (e.g., "FirstLast_A02_CodingBasics.Rmd") prior to submission.



## Basics Day 1

1. Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name. 

2. Compute the mean and median of this sequence. 

3. Ask R to determine whether the mean is greater than the median. 

4. Insert comments in your code to describe what you are doing.

```{r}
#1.
four_sequence <- seq(1,100,4) #creating sequence from 1 to 100 by 4
four_sequence #printing sequence

#2. 
mean_seq <- mean(four_sequence) #calculating mean of four_sequence
mean_seq

median_seq <- median(four_sequence) #calculating median of four_sequence
median_seq

#3. 
mean_seq > median_seq #checking whether mean is greater than median


```

## Basics Day 2

5. Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6. Label each vector with a comment on what type of vector it is.

7. Combine each of the vectors into a data frame. Assign the data frame an informative name.

8. Label the columns of your data frame with informative titles.

```{r}
#a
students <- c("Stephanie", "Asher", "Alex", "Nicole") #creates vector of student names
typeof(students)
#students = character vector

#b
test_scores <- c(100, 99, 48, 87) #creates vector of test scores
typeof(test_scores)
#test_scores = double vector

#c
passing_grade <- c(TRUE, TRUE, FALSE, TRUE) #creates vector of grades that are passing = TRUE or failed = FALSE
typeof(passing_grade)
#passing_grade = logical vector

student_tests_df <- data.frame("Student" = students, "Test_Score" = test_scores, "Passing_Grade" = passing_grade)
student_tests_df


```

9. QUESTION: How is this data frame different from a matrix?

> Answer: A data frame can include different types of data, such as character, logical, or numeric vectors.  However, a matrix can only include a single type of data. This data frame contains three types of vectors: character, double, and logical.

10. Create a function with an if/else statement. Your function should determine whether a test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. Hint: Use `print`, not `return`. The name of your function should be informative.

11. Apply your function to the vector with test scores that you created in number 5. 

```{r}
grade_pass <- function(x){
  ifelse(x>=50, print(TRUE), print(FALSE))
}

grade_pass(test_scores)

```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: Ifelse worked because it applies the function to all elements in a vector, whereas if and else only apply the function to the first element.
