---
title: "Assignment 5: Data Visualization"
author: "Stephanie Kinser"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_A05_DataVisualization.Rmd") prior to submission.

The completed exercise is due on Monday, February 14 at 7:00 pm.



## Set up your session 
1. Set up your session. Verify your working directory and load the tidyverse 
and cowplot packages. Upload the NTL-LTER processed data files for nutrients 
and chemistry/physics for Peter and Paul Lakes (use the tidy 
[`NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv`] version) and 
the processed data file for the Niwot Ridge litter dataset (use the [`NEON_NIWO_Litter_mass_trap_Processed.csv`] version). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r}
#1 
getwd()
#install.packages("tidyverse")
#install.packages("cowplot")
library(tidyverse)
library(cowplot)

NTL_LTER <- read.csv("./Data/Processed/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv")

Litter <- read.csv("./Data/Processed/NEON_NIWO_Litter_mass_trap_Processed.csv")

NTL_LTER$sampledate <- as.Date(NTL_LTER$sampledate, format = "%Y-%m-%d")

Litter$collectDate <- as.Date(Litter$collectDate, format = "%Y-%m-%d")

#2 
NTL_LTER$sampledate <- as.Date(NTL_LTER$sampledate, format = "%Y-%m-%d")
class(NTL_LTER$sampledate)

Litter$collectDate <- as.Date(Litter$collectDate, format = "%Y-%m-%d")
class(Litter$collectDate)

```
## Define your theme
3. Build a theme and set it as your default theme.

```{r}
#3
A5_theme <- theme_light(base_size = 12)+
   theme(axis.text = element_text(color = "black"), 
        legend.position = "right", panel.grid.major = element_blank(), panel.grid.minor = element_bl)

theme_set(A5_theme)

```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add a line of best fit and color it black. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and `ylim()`).

```{r}
#4
#install.packages("ggridges")
library(ggridges)

plot4 <- ggplot(NTL_LTER, aes(x = po4, y = tp_ug)) +
  geom_point(color = "darkgreen")+
  facet_wrap(vars(lakename), nrow = 2)+
  xlim(0,50)+
  geom_smooth(method = lm, color = "black")+
  labs(x = "Phosphate Concentration (ug)", y = "Phosphorus Concentration (ug)") #check units***
print(plot4)

```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned.

```{r}
#this section needs a lot of work:
  #check axes spacing and labeling and consolidate so that there are fewer labels
#I don't get what the x axis is showing.

#5
NTL_temp <- ggplot(NTL_LTER)+
  geom_boxplot(aes(x = month, y = temperature_C, color = lakename))+
  labs(x = "Month", y = "Temperature (Celsius)")
print(NTL_temp)

#What is TP and TN
NTL_TP <- ggplot(NTL_LTER)+
  geom_boxplot(aes(x = month, y = tp_ug, color = lakename))+
  labs(x = "Month", y = "Phosphorus Concentration (ug)") #check units
print(NTL_TP)

NTL_TN <- ggplot(NTL_LTER)+
  geom_boxplot(aes(x = month, y = tn_ug, color = lakename))+
  scale_x_date(date_breaks = "1 month")+
  labs(x = "Month", y = "Nitrogen Concentration (ug)") #check units
print(NTL_TN)

NTL_grid <- plot_grid(NTL_temp, NTL_TP, NTL_TN, nrow = 3, #says to plot one below the other
          align = 'h')
          #rel_heights = c(1.25, 1)) #adjust how much space each plot need and have cells of different sizes
print(NTL_grid)

```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: 

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r}
#6
Needles <- ggplot(subset(Litter, functionalGroup =="Needles"))+
  geom_point(aes(x = collectDate, y = dryMass, color = nlcdClass))+
  scale_x_date(date_breaks = "6 months", date_labels = "%m-%Y")+
  labs(x = "Date", y = "Dry Mass (g)", color = "Class")+ #check units for dry mass
  scale_color_hue(labels = c("Evergreen/Forest", "Grassland/Herbaceous", "Shrub/Scrub")) #adjusting labels in legend
print(Needles)


#7
ncld_labels <- as_labeller(c(evergreenForest = "Evergreen/Forest", grasslandHerbaceous = "Grassland/Herbaceous", shrubScrub = "Shrub/Scrub")) #creates labels for facets - applied with labeller

Needles_facet <- ggplot(subset(Litter, functionalGroup =="Needles"))+
  geom_point(aes(x = collectDate, y = dryMass))+
  facet_wrap(vars(nlcdClass), nrow = 3, labeller = ncld_labels)+
  scale_x_date(date_breaks = "6 months", date_labels = "%m-%Y")+
  labs(x = "Date", y = "Dry Mass (g)") #check units for dry mass
print(Needles_facet)


```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: Plot 7 is more effective because you can more clearly see the individual points for each class and, thereby, compare the differences in classes more easily.
