---
title: "Assignment 09: Data Scraping"
author: "Stephanie Kinser"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

# Total points:

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_09_Data_Scraping.Rmd") prior to submission.


## Set up 
1. Set up your session:

* Check your working directory
* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Set your ggplot theme

```{r, message = FALSE}
#1

getwd()

library(tidyverse)
library(dplyr)
#install.packages("lubridate")
library(lubridate)
#install.packages("rvest")
library(rvest)

A9_theme <- theme_light(base_size = 12)+
   theme(axis.text = element_text(color = "black"), 
        legend.position = "right", panel.grid.minor = element_blank())

theme_set(A9_theme)

```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2019 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Change the date from 2021 to 2020 in the upper right corner.
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2020>

Indicate this website as the as the URL to be scraped. (In other words, read the contents into an
`rvest` webpage object.)

```{r set.the.scraping.website}
#2
webpage <- read_html('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2020')

webpage

```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PSWID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Average Daily Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to three separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values, with the first value being 36.0100.

```{r scrape.the.data}
#3
water.system.name <- webpage %>% html_nodes('div+ table tr:nth-child(1) td:nth-child(2)') %>% html_text()

pwsid <- webpage %>% html_nodes('td tr:nth-child(1) td:nth-child(5)') %>% html_text()

ownership <- webpage %>% html_nodes('div+ table tr:nth-child(2) td:nth-child(4)') %>% html_text()

max.withdrawals.mgd <- webpage %>% html_nodes('th~ td+ td') %>% html_text()

```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in date format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in order. You can overcome this by creating a month column in the same order the data are scraped: Jan, May, Sept, Feb, etc...

5. Plot the max daily withdrawals across the months for 2020

```{r create.a.dataframe.from.scraped.data}

#4
Month <- c(01, 05, 09, 02, 06, 10, 03, 07, 11, 04, 08, 12)
Year <- rep(2020, 12)

water_df <- data.frame(
  "Month" = Month,
  "Year" = Year,
  "Date" = my(paste(Month,"-",Year)),
  "Water_System" = rep(water.system.name, 12),
  "Owner" = rep(ownership, 12),
  "PWSID" = rep(pwsid, 12),
  "Max_Withdrawals" = round(as.numeric(max.withdrawals.mgd), 2)) %>%
  arrange(Date)


#5
ggplot(water_df, aes(x = Date, y = Max_Withdrawals)) +
  geom_point() +
  labs(title = "Durham Max Water Withdrawals (2020)", x = "Date", y = "Maximum Daily Water Withdrawal (MGD)")


```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function using your code above that can scrape data for any PWSID and year for which the NC DEQ has data. **Be sure to modify the code to reflect the year and site scraped**.

```{r construct.a.scraping.function}
#6.
scrape.it <- function(PWSID, the_year){
  
  #Get the proper webpage
    webpage <- read_html(paste0('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=', PWSID, '&year=', the_year))
    

  #Locate elements and read their text attributes into variables
  water.system.name <- webpage %>% html_nodes('div+ table tr:nth-child(1) td:nth-child(2)') %>% html_text()

  pwsid <- webpage %>% html_nodes('td tr:nth-child(1) td:nth-child(5)') %>% html_text()

  ownership <- webpage %>% html_nodes('div+ table tr:nth-child(2) td:nth-child(4)') %>% html_text()

  max.withdrawals.mgd <- webpage %>% html_nodes('th~ td+ td') %>% html_text()

  
  #Construct a dataframe from the values
  Month <- c(01, 05, 09, 02, 06, 10, 03, 07, 11, 04, 08, 12)
  Year <- rep(the_year, 12)

  the_df <- data.frame(
    "Month" = Month,
    "Year" = Year,
    "Date" = my(paste(Month,"-",Year)),
    "Water_System" = rep(water.system.name, 12),
    "Owner" = rep(ownership, 12),
    "PWSID" = rep(pwsid, 12),
    "Max_Withdrawals" = round(as.numeric(max.withdrawals.mgd), 2)) %>%
    arrange(Date)
  
  return(the_df)
}


```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2015

```{r fetch.and.plot.Durham.2015.data}
#7
Durham_2015 <- scrape.it("03-32-010", 2015)

ggplot(Durham_2015, aes(x = Date, y = Max_Withdrawals)) +
  geom_point() +
  labs(title = "Durham Max Water Withdrawals (2015)", x = "Date", y = "Maximum Daily Water Withdrawal (MGD)")

```

8. Use the function above to extract data for Asheville (PWSID = 01-11-010) in 2015. Combine this data with the Durham data collected above and create a plot that compares the Asheville to Durham's water withdrawals.

```{r fetch.and.plot.Durham.2015.data}
#8
Asheville_2015 <- scrape.it("01-11-010", 2015)

water_2015 <- rbind(Durham_2015, Asheville_2015) %>%
  group_by(Water_System)

water_plot <- ggplot(water_2015)+
  geom_point(aes(x = Date, y = Max_Withdrawals, color = Water_System))+
  labs(title = "Monthly Water Use", x = "Month", y = "Maximum Daily Water Withdrawal (MGD)", color = "Water System")

water_plot

```


9. Use the code & function you created above to plot Asheville's max daily withdrawal by months for the years 2010 thru 2019.Add a smoothed line to the plot.

```{r}
#9
Asheville_2010 <- scrape.it("01-11-010", 2010)
Asheville_2011 <- scrape.it("01-11-010", 2011)
Asheville_2012 <- scrape.it("01-11-010", 2012)
Asheville_2013 <- scrape.it("01-11-010", 2013)
Asheville_2014 <- scrape.it("01-11-010", 2014)
Asheville_2015 <- scrape.it("01-11-010", 2015)
Asheville_2016 <- scrape.it("01-11-010", 2016)
Asheville_2017 <- scrape.it("01-11-010", 2017)
Asheville_2018 <- scrape.it("01-11-010", 2018)
Asheville_2019 <- scrape.it("01-11-010", 2019)

Asheville_water <- rbind(Asheville_2010, Asheville_2011, Asheville_2012, Asheville_2013, Asheville_2014, Asheville_2015, Asheville_2016, Asheville_2017, Asheville_2018, Asheville_2019)

Asheville_plot <- ggplot(Asheville_water, aes(x = Date, y = Max_Withdrawals)) +
  geom_point()+
  geom_smooth(method = lm, se = FALSE)+
  labs(title = "Asheville Maximum Water Withdrawals", x = "Date", y = "Maximum Daily Water Withdrawal (MGD)")

Asheville_plot
```

>Question: Just by looking at the plot (i.e. not running statistics), does Asheville have a trend in water usage over time?
The plot suggests that there has been an increase in maximum water usage from 2010 through 2019.